---
title: "visualization_colorectal"
format:
  html:
    toc: true
    code-fold: true
editor: source
execute:
  echo: true
  warning: false
  message: false
  error: false
---

```{r setup,message=FALSE,warning=FALSE}
library(dplyr)
library(readr)
#install.packages("janitor")
library(janitor)
library(ggplot2)
```

# Load Raw Export

```{r}
data_cr_cancer <- read_csv(
  "/Users/apple/Desktop/me/WCM/semester/research/qiu/smoke/cancer_main_with_RPL.csv",
  show_col_types = FALSE
)

df <- read_csv("/Users/apple/Desktop/me/WCM/semester/research/qiu/smoke/cancer_main_with_RPL.csv")
```

# Define Colorectal Cancer Cohort

```{r warning=FALSE,message=FALSE}
data_cr_cancer <- data_cr_cancer %>%
  clean_names()

crc_sites <- c(
  "Ascending Colon",
  "Cecum",
  "Descending Colon",
  "Hepatic Flexure",
  "Sigmoid Colon",
  "Splenic Flexure",
  "Transverse Colon",
  "Large Intestine, NOS",
  "Rectosigmoid Junction",
  "Rectum"
)

crc_dat <- data_cr_cancer %>%
  filter(site_recode_icd_o_3_who_2008 %in% crc_sites)
```




```{r warning=FALSE,message=FALSE}


```



# Recode Variables

```{r echo=FALSE}
crc_dat <- crc_dat %>%
  mutate(
    age_group_recode = case_when(
      age_recode_with_1_year_olds %in% c(
        "00 years", "01-04 years", "05-09 years", "10-14 years", "15-19 years",
        "20-24 years", "25-29 years"
      ) ~ "<30",
      age_recode_with_1_year_olds %in% c(
        "30-34 years", "35-39 years"
      ) ~ "30-39",
      age_recode_with_1_year_olds %in% c(
        "40-44 years", "45-49 years"
      ) ~ "40-49",
      age_recode_with_1_year_olds %in% c(
        "50-54 years", "55-59 years", "60-64 years"
      ) ~ "50-64",
      age_recode_with_1_year_olds %in% c(
        "65-69 years", "70-74 years", "75-79 years"
      ) ~ "65-79",
      age_recode_with_1_year_olds %in% c(
        "80-84 years", "85+ years"
      ) ~ "80+",
      TRUE ~ NA_character_
    )
  )
```


```{r}
crc_dat <- crc_dat %>%
  filter(year_of_diagnosis >= 2010 & year_of_diagnosis <= 2022)
#summary(crc_dat)
```




```{r}
crc_dat <- crc_dat %>%
  mutate(
    age_under50 = case_when(
      age_group_recode %in% c("<30", "30-39", "40-49") ~ 1,
      age_group_recode %in% c("50-64", "65-79", "80+") ~ 0,
      TRUE ~ NA_real_
    )
  )
```


```{r warning=FALSE,message=FALSE}
library(gtsummary)

table_overall <- crc_dat %>%
  select(
    age_recode_with_1_year_olds,
    sex,
    race_recode_white_black_other,
    year_of_diagnosis,
    origin_recode_nhia_hispanic_non_hisp,
    rpl_themes
  ) %>%
  tbl_summary(
    statistic = list(
      all_categorical() ~ "{n} ({p}%)",
      rpl_themes ~ "{median} ({p25}, {p75})"
    ),
    digits = rpl_themes ~ 3,
    missing = "no",#ifany
    label = list(
      age_recode_with_1_year_olds ~ "Age",
      sex ~ "Sex",
      race_recode_white_black_other ~ "Race",
      year_of_diagnosis ~ "Year of Diagnosis",
      origin_recode_nhia_hispanic_non_hisp ~ "Hispanic Origin",
      rpl_themes ~ "Overall Social Vulnerability (RPL_THEMES)"
    )
  ) %>%
  bold_labels() %>%
  modify_caption(
    "**Table 1. Descriptive raw data statistics with overall column distribution**"
  )

table_overall
```


```{r warning=FALSE,message=FALSE}

table_under50 <- crc_dat %>%
  filter(age_under50 == 1) %>%
  select(
    age_recode_with_1_year_olds,
    sex,
    race_recode_white_black_other,
    year_of_diagnosis,
    origin_recode_nhia_hispanic_non_hisp,
    rpl_themes
  ) %>%
  tbl_summary(
    statistic = list(
      all_categorical() ~ "{n} ({p}%)",
      rpl_themes ~ "{median} ({p25}, {p75})"
    ),
    digits = rpl_themes ~ 3,
    missing = "ifany",
    label = list(
      age_recode_with_1_year_olds ~ "Age",
      sex ~ "Sex",
      race_recode_white_black_other ~ "Race",
      year_of_diagnosis ~ "Year of Diagnosis",
      origin_recode_nhia_hispanic_non_hisp ~ "Hispanic Origin",
      rpl_themes ~ "Overall Social Vulnerability (RPL_THEMES)"
    )
  ) %>%
  bold_labels() %>%
  modify_caption(
    "**Table 2. Descriptive raw data statistics among colorectal cancer cases aged <50 years**"
  )

table_under50

```



### <50 CRC overview

```{r}
trend_overall <- crc_dat %>%
  group_by(year_of_diagnosis) %>%
  summarise(
    total_n = n(),
    under50_n = sum(age_under50, na.rm = TRUE),
    pct_under50 = under50_n / total_n * 100
  )
```



```{r}


ggplot(trend_overall, aes(x = year_of_diagnosis, y = pct_under50)) +
  geom_line(size = 1) +
  geom_point() +
  scale_x_continuous(breaks = 2010:2022) +
  labs(
    title = "Trend of <50 Colorectal Cancer Patients (2010–2022)",
    x = "Year",
    y = "Percentage (%)"
  ) +
  theme_minimal()
```


##### CRC Percentage by Under-50 Age Group


```{r}

trend_agegroup_pct <- crc_dat %>%
  filter(age_group_recode %in% c("<30", "30-39", "40-49")) %>%
  group_by(year_of_diagnosis, age_group_recode) %>%
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  left_join(
    crc_dat %>%
      group_by(year_of_diagnosis) %>%
      summarise(total_crc = n(), .groups = "drop"),
    by = "year_of_diagnosis"
  ) %>%
  mutate(
    pct_of_all_crc = n / total_crc * 100
  )

ggplot(trend_agegroup_pct, aes(x = year_of_diagnosis, y = pct_of_all_crc, color = age_group_recode)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = 2010:2022) +
  labs(
    title = "Trend of CRC Percentage by Under-50 Age Group (2010-2022)",
    x = "Year",
    y = "Percentage of all CRC patients (%)",
    color = "Age group"
  ) +
  theme_minimal()
```




### <50 CRC by sex


```{r}
trend_sex <- crc_dat %>%
  group_by(year_of_diagnosis, sex) %>%
  summarise(
    total_n = sum(!is.na(age_under50)),
    under50_n = sum(age_under50, na.rm = TRUE),
    pct_under50 = under50_n / total_n * 100,
    .groups = "drop"
  )
```

```{r}
ggplot(trend_sex, aes(x = year_of_diagnosis, y = pct_under50, color = sex)) +
  geom_line(size = 1) +
  geom_point() +
  scale_x_continuous(breaks = 2010:2022) +
  labs(
    title = "Trend of <50 CRC by Sex",
    x = "Year",
    y = "Percentage (%)",
    color = "Sex"
  ) +
  theme_minimal()
```


##### stacked composition

```{r echo=FALSE,include=FALSE}
trend_under50_age_by_sex <- crc_dat %>%
  filter(age_group_recode %in% c("<30", "30-39", "40-49")) %>%
  filter(!is.na(sex)) %>%
  group_by(year_of_diagnosis, sex, age_group_recode) %>%
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    age_group_recode = factor(age_group_recode, levels = c("<30", "30-39", "40-49"))
  )
ggplot(trend_under50_age_by_sex,
       aes(x = factor(year_of_diagnosis), y = n, fill = age_group_recode)) +
  geom_col() +
  facet_wrap(~ sex) +
  labs(
    title = "Under-50 CRC Cases by Age Group and Sex",
    x = "Year",
    y = "Number of patients",
    fill = "Age group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )
```





### <50 CRC by race

```{r}
crc_dat <- crc_dat %>%
  mutate(
    race_simple = case_when(
      race_recode_white_black_other == "Other (American Indian/AK Native, Asian/Pacific Islander)" ~ "Other",
      race_recode_white_black_other == "Unknown (including other unspec 1991+)" ~ "Unknown",
      TRUE ~ race_recode_white_black_other
    )
  )

trend_race <- crc_dat %>%
  filter(!is.na(race_simple)) %>%
  group_by(year_of_diagnosis, race_simple) %>%
  summarise(
    total_n = sum(!is.na(age_under50)),
    under50_n = sum(age_under50, na.rm = TRUE),
    pct_under50 = under50_n / total_n * 100,
    .groups = "drop"
  )
```



```{r}

ggplot(trend_race, aes(x = year_of_diagnosis, y = pct_under50, color = race_simple)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  
  scale_x_continuous(breaks = 2010:2022) +
  
  labs(
    title = "Trend of <50 CRC by Race (2010–2022)",
    x = "Year",
    y = "Percentage (%)",
    color = "Race"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```



##### stacked composition


```{r echo=FALSE,include=FALSE}
trend_under50_age_by_race <- crc_dat %>%
  filter(age_group_recode %in% c("<30", "30-39", "40-49")) %>%
  filter(!is.na(race_simple)) %>%
  group_by(year_of_diagnosis, race_simple, age_group_recode) %>%
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    age_group_recode = factor(
      age_group_recode,
      levels = c("<30", "30-39", "40-49")
    )
  )

ggplot(trend_under50_age_by_race,
       aes(x = year_of_diagnosis, y = n, fill = age_group_recode)) +
  
  geom_col() +
  
  facet_wrap(~ race_simple) +
  
  scale_x_continuous(breaks = 2010:2022) +
  
  scale_fill_manual(values = c(
    "<30" = "#E64B35",
    "30-39" = "#4DBBD5",
    "40-49" = "#00A087"
  )) +
  
  labs(
    title = "Under-50 CRC Cases by Age Group and Race",
    x = "Year",
    y = "Number of patients",
    fill = "Age group"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )

```



### <50 CRC by Social Vulnerability Index

```{r}

#ggplot(
#  crc_dat %>% filter(age_under50 == 1),
#  aes(
#    x = factor(year_of_diagnosis),
#    y = rpl_themes
#  )
#) +
#  geom_boxplot() +
#  labs(
#    title = "Distribution of Overall RPL_THEMES Among CRC Cases Aged <50",
#    x = "Year of Diagnosis",
#    y = "RPL_THEMES"
#  ) +
#  theme_minimal(base_size = 14) +
#  theme(
#    plot.title = element_text(face = "bold")
#  )

ggplot(
  crc_dat %>% 
    filter(age_under50 == 1),
  aes(
    x = year_of_diagnosis,
    y = rpl_themes
  )
) +
  
  geom_boxplot(
    aes(group = year_of_diagnosis),
    width = 0.6,
    outlier.alpha = 0.25
  ) +
  
  stat_summary(
    fun = median,
    geom = "line",
    linewidth = 1.2
  ) +
  
  stat_summary(
    fun = median,
    geom = "point",
    size = 2.5
  ) +
  
  # LOESS
  geom_smooth(
    method = "loess",
    se = FALSE,
    linewidth = 1,
    linetype = 2
  ) +
  
  scale_x_continuous(
    breaks = 2010:2022
  ) +
  
  labs(
    title = "RPL_THEMES Distribution Among CRC Cases Aged <50",
    subtitle = "Boxplots with median trend across diagnosis years",
    x = "Year of Diagnosis",
    y = "RPL_THEMES"
  ) +
  
  theme_classic(base_size = 14) +
  
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```



```{r include=FALSE}

ggplot(
  crc_dat %>% filter(age_under50 == 1),
  aes(
    x = factor(year_of_diagnosis),
    y = rpl_themes
  )
) +
  geom_violin(trim = FALSE) +
  labs(
    title = "Distribution of RPL_THEMES Among CRC Cases Aged <50",
    x = "Year",
    y = "RPL_THEMES"
  ) +
  theme_minimal()

```


```{r echo=FALSE,include=FALSE}
crc_dat %>%
  group_by(age_under50) %>%
  summarise(
    n = n(),
    mean_rpl = mean(rpl_themes, na.rm = TRUE),
    sd_rpl = sd(rpl_themes, na.rm = TRUE),
    median_rpl = median(rpl_themes, na.rm = TRUE),
    p25 = quantile(rpl_themes, 0.25, na.rm = TRUE),
    p75 = quantile(rpl_themes, 0.75, na.rm = TRUE)
  )

```
```{r}

t.test(
  rpl_themes ~ age_under50,
  data = crc_dat
)


wilcox.test(
  rpl_themes ~ age_under50,
  data = crc_dat
)
```

<50 CRC patients have significantly HIGHER social vulnerability.


### <50 CRC by Hispanic
```{r echo=FALSE,include=FALSE}

crc_dat %>%
  group_by(origin_recode_nhia_hispanic_non_hisp) %>%
  summarise(
    n = n(),
    mean_rpl = mean(rpl_themes, na.rm = TRUE),
    median_rpl = median(rpl_themes, na.rm = TRUE)
  )


```


```{r}

crc_under50 <- crc_dat %>%
  filter(age_under50 == 1)

wilcox.test(
  rpl_themes ~ origin_recode_nhia_hispanic_non_hisp,
  data = crc_under50
)

```
Among <50 CRC，Hispanic is more vulnerable.


```{r}
#========================================================
# Trend of Hispanic proportion among CRC cases aged <50
#========================================================

trend_hispanic_under50 <- crc_dat %>%
  filter(age_under50 == 1) %>%
  mutate(
    hispanic_group = case_when(
      origin_recode_nhia_hispanic_non_hisp == "Spanish-Hispanic-Latino" ~ "Hispanic",
      origin_recode_nhia_hispanic_non_hisp == "Non-Spanish-Hispanic-Latino" ~ "Non-Hispanic",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(hispanic_group)) %>%
  group_by(year_of_diagnosis, hispanic_group) %>%
  summarise(
    n = n(),
    .groups = "drop"
  ) %>%
  left_join(
    crc_dat %>%
      filter(age_under50 == 1) %>%
      group_by(year_of_diagnosis) %>%
      summarise(
        total_under50 = n(),
        .groups = "drop"
      ),
    by = "year_of_diagnosis"
  ) %>%
  mutate(
    pct = n / total_under50 * 100
  )

ggplot(
  trend_hispanic_under50,
  aes(
    x = year_of_diagnosis,
    y = pct,
    color = hispanic_group
  )
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = 2010:2022
  ) +
  labs(
    title = "Trend of Hispanic Proportion Among CRC Cases Aged <50",
    x = "Year of Diagnosis",
    y = "Percentage (%)",
    color = "Group"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )
```


